---
title: "All of Us Phenotype Selection"
author: "Katie Hooker"
date: "2025-01-22"
output: html_document
---

The All of Us phenotypes index was downloaded [here](https://support.researchallofus.org/hc/en-us/articles/27049847988884-Overview-of-the-All-by-All-tables-available-on-the-All-of-Us-Researcher-Workbench#01HZ2KKW3XZ55MSBT7G1RNQS86) for identifying relevant phenotypes for subsequent analysis. The Excel file was converted to a .csv for ease of use.

# Reading in the index
```{r}
suppressMessages(library(dplyr))
pheno_index <- read.csv("./data/AoU/All_by_All_phenotypes.csv")

table(pheno_index$category)
table(pheno_index$phecode_category)

```
# Filtering for European ancestry
Because the LD blocks used in S-LDSC are derived from European ancestry, we will restrict our analysis to this group

```{r}
pheno_index_EUR <- pheno_index %>% subset(ancestry == "eur")

```

# Filtering for sample size $ >= 10,000$

```{r}
pheno_index_EUR <- pheno_index_EUR %>% mutate(n_total = n_cases + n_controls)

pheno_index_10k <- pheno_index_EUR %>% subset(n_total >= 10000)
table(pheno_index_10k$phecode_category)

```
# Filtering for phecodeX rather than phecode because it is more granular
https://academic.oup.com/bioinformatics/article/39/11/btad655/7335839\
We cannot use both phecodeX and phecode as they are both classification schemes and there will be overlap in phenotypes?

```{r}
table(pheno_index_10k$category)
pheno_index_filtered <- pheno_index_10k %>% filter(category == "phecodeX")

```

# Filtering for digestive, gastrointestinal, and neoplasms phecode_category

```{r}
table(pheno_index_filtered$phecode_category)
pheno_index_filtered <- pheno_index_filtered %>% filter(phecode_category == 'Gastrointestinal' |
                                                phecode_category == 'Neoplasms')
# Should I include endocrine/metabolic, genetic, congenital, or null??

```

# Removing irrelevant phenotypes by keyword

```{r}
library(stringr)
pheno_index_filtered <- pheno_index_filtered %>% filter(# sex-specific
                                                          str_detect(description, "(?i)breast", negate=TRUE) &
                                                          str_detect(description, "(?i)cervi", negate=TRUE) &
                                                          str_detect(description, "(?i)uter", negate=TRUE) &
                                                          str_detect(description, "(?i)ovar", negate=TRUE) &
                                                          str_detect(description, "(?i)gyneco", negate=TRUE) &
                                                          str_detect(description, "(?i)mammo", negate=TRUE) &
                                                          str_detect(description, "(?i)endometr", negate=TRUE) &
                                                          str_detect(description, "(?i)prostate", negate=TRUE) &
                                                          str_detect(description, "(?i)genit", negate=TRUE) &
                                                          # respiratory
                                                          str_detect(description, "(?i)respiratory", negate=TRUE) &
                                                          str_detect(description, "(?i)lung", negate=TRUE) &
                                                          # dermatological
                                                          str_detect(description, " skin", negate=TRUE) &
                                                          str_detect(description, "carcin", negate=TRUE) &
                                                          # genitourinary
                                                          str_detect(description, " bladder", negate=TRUE) &
                                                          # musculoskeletal
                                                          str_detect(description, "(?i)myelo", negate=TRUE) &
                                                          str_detect(description, "(?i)sarcoma", negate=TRUE) &
                                                          str_detect(description, "(?i)connective", negate=TRUE) &
                                                          # neuro
                                                          str_detect(description, "(?i)brain", negate=TRUE) &
                                                          # lymphatic
                                                          str_detect(description, "(?i)leukemia", negate=TRUE) &
                                                          str_detect(description, "(?i)lymph", negate=TRUE) &
                                                          # eye
                                                          str_detect(description, "(?i)eye", negate=TRUE) &
                                                          # findings
                                                          str_detect(description, "(?i)status", negate=TRUE) &
                                                          str_detect(description, "(?i)findings", negate=TRUE) &
                                                          # mouth, teeth, head, etc
                                                          str_detect(description, "(?i)lip", negate=TRUE) &
                                                          str_detect(description, "(?i)teeth", negate=TRUE) &
                                                          str_detect(description, "(?i)tooth", negate=TRUE) &
                                                          str_detect(description, "(?i)dent", negate=TRUE) &
                                                          str_detect(description, "(?i)mouth", negate=TRUE) &
                                                          str_detect(description, "(?i)gingi", negate=TRUE) &
                                                          str_detect(description, "(?i)head", negate=TRUE) &
                                                          str_detect(description, "(?i)oral", negate=TRUE) &
                                                          str_detect(description, "(?i)nasal", negate=TRUE) &
                                                          str_detect(description, "(?i)salivary", negate=TRUE) &
                                                          str_detect(description, "(?i)peria", negate=TRUE) &
                                                          str_detect(description, "(?i)perio", negate=TRUE) &
                                                          # endocrine, neuroendocrine, thyroid
                                                          str_detect(description, "(?i)endocrine", negate=TRUE) &
                                                          str_detect(description, "(?i)thyroid", negate=TRUE) &
                                                          # not specific
                                                          str_detect(description, "(?i)unspecified", negate=TRUE) &
                                                            str_detect(description, "(?i)nonspecific", negate=TRUE) &
                                                          str_detect(description, "(?i)other", negate=TRUE))

```

Kidney? Endocrine? Thyroid?

# Removing any last irrelevant phenotypes on a specific basis
```{r}
pheno_index_filtered <- pheno_index_filtered %>% filter(description != "Carcinoid tumors" &
                                                          description != "Pheochromocytoma" &
                                                          description != "Secondary malignant neoplasm" &
                                                          description != "Carcinoid tumors" &
                                                          description != "Hemo onc - by cell of origin" &
                                                          description != "Mature B-cell" &
                                                          description != "Essential thrombocythemia" &
                                                          description != "Sequelae of cancer" &
                                                          description != "Essential thrombocythemia" &
                                                          description != "Essential thrombocythemia" &
                                                          description != "Essential thrombocythemia" &
                                                          description != "Intestinal infection" &
                                                          description != "Bacterial enteritis" &
                                                          description != "Intestinal infection due to C. difficile" &
                                                          description != "Nausea and vomiting" &
                                                          description != "Nausea" &
                                                          description != "Vomiting" &
                                                          description != "Symptoms involving digestive system" &
                                                          description != "Diarrhea" &
                                                          description != "Flatulence, eructation, and gas pain" &
                                                          description != "Nevus, non-neoplastic" &
                                                          description != "Cancer (solid tumor, excluding BCC)" &
                                                          description != "Melanocytic nevi*" &
                                                          description != "Benign neoplasm of peripheral nerves*" &
                                                          description != "Hemangioma" &
                                                          description != "Benign neoplasm of choroid" &
                                                          description != "Benign neoplasm of meninges (Meningioma)" &
                                                          description != "Benign neoplasm of the pituitary gland and craniopharyngeal duct" &
                                                          description != "Sialoadenitis" &
                                                          description != "Stomatitis" &
                                                          description != "Diseases of tongue" &
                                                          description != "Glossitis")

for (i in 1:nrow(pheno_index_filtered)) {
  print(pheno_index_filtered[i,]$description)
  print(pheno_index_filtered[i,]$n_cases)
  print(pheno_index_filtered[i,]$n_total)
}

```

# Printing a list of phecodes for data pulling in the AoU Researcher Workbench

```{r}
phecodes <- pheno_index_filtered$phenoname

python_format <- paste0("[", paste(sapply(phecodes, function(code) paste0('"', code, '"')), collapse = ", "), "]")

cat(python_format)

```

We pivoted from All of Us because after contacting their research support team, I was informed that the All by All GWAS summary statistics are no longer being approved for download.
