---
title: "MVP Phenotype Selection"
author: "Katie Hooker"
date: "2025-01-22"
output: html_document
---

# Reading in the phenotype manifest
Downloaded from https://www-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/projects/gap/cgi-bin/study.cgi?study_id=phs002453.v1.p1
https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/

```{r}
library(dplyr)
pheno_mt <- read.csv("./data/phs002453.v1-Documents/MVP_R4.1000G_AGR.DataDictionary_and_Counts_vJuly2024-1.csv")
names(pheno_mt)

pheno_mt <- pheno_mt %>% select(Trait, Trait_Type, Category, SubCategory, Description, num_samples.EUR, num_cases.EUR, num_controls.EUR, dbgap_cat.EUR, dbgap_cat.pheno)

```

# Filter for 10k cases if binary trait and 10k total cohort if continous trait

```{r}
pheno_filt <- pheno_mt %>%
  filter(
    (Trait_Type == "binary" & num_cases.EUR > 10000) |
    (Trait_Type == "qt" & num_samples.EUR > 10000)
  )

table(pheno_filt$Category)

```

# Filter for category of labs, vitals, and phecodes

```{r}
pheno_filt <- pheno_filt %>% filter(Category == "Labs" |
                                      Category == "PheCodes" |
                                      Category == "Vitals")
```

# Filter by subcategory for phecodes

```{r}
table(pheno_filt$SubCategory)

pheno_filt <- pheno_filt %>% filter(SubCategory == "Labs" |
                                      SubCategory == "Vitals" |
                                      SubCategory == "Digestive System" |
                                      SubCategory == "Endocrine or Metabolic" |
                                      SubCategory == "Neoplasms" |
                                      SubCategory == "Symptoms")
```
# Removing irrelevant phenotypes by keyword

```{r}
library(stringr)
pheno_filt <- pheno_filt %>% filter(str_detect(Description, "(?i)cervi", negate=TRUE) &
                                      str_detect(Description, "(?i)prostate", negate=TRUE) &
                                      str_detect(Description, "(?i)testicu", negate=TRUE) &
                                      str_detect(Description, "(?i)bladder", negate=TRUE) &
                                      # respiratory
                                      str_detect(Description, "(?i)respiratory", negate=TRUE) &
                                      str_detect(Description, "(?i)lung", negate=TRUE) &
                                      # dermatological
                                      str_detect(Description, "(?i)skin", negate=TRUE) &
                                      str_detect(Description, "carcin", negate=TRUE) &
                                      # genitourinary
                                      str_detect(Description, "(?i)bladder", negate=TRUE) &
                                      # neuro
                                      str_detect(Description, "(?i)brain", negate=TRUE) &
                                      # lymphatic
                                      str_detect(Description, "(?i)lymph", negate=TRUE) &
                                      # musculoskeletal
                                      str_detect(Description, "(?i)muscle", negate=TRUE) &
                                      str_detect(Description, "(?i)limb", negate=TRUE) &
                                      # eye
                                      str_detect(Description, "(?i)eye", negate=TRUE) &
                                      # mouth, teeth, head, etc
                                      str_detect(Description, "(?i)teeth", negate=TRUE) &
                                      str_detect(Description, "(?i)tooth", negate=TRUE) &
                                      str_detect(Description, "(?i)dent", negate=TRUE) &
                                      str_detect(Description, "(?i)gingi", negate=TRUE) &
                                      str_detect(Description, "(?i)oral", negate=TRUE) &
                                      str_detect(Description, "(?i)peria", negate=TRUE) &
                                      str_detect(Description, "(?i)perio", negate=TRUE) &
                                      # screenings, findings, etc
                                      str_detect(Description, "(?i)screening", negate=TRUE) &
                                      str_detect(Description, "(?i)therapy", negate=TRUE))
```


# If the category is a lab or vital measure, keep only the mean value
```{r}
pheno_filt <- pheno_filt %>%
  filter((Category == "Labs" & grepl("Mean", Trait)) | 
           (Category == "Vitals" & grepl("Mean", Trait)) |
           Category == "PheCodes")

```

# Further filtering on a case by case basis
```{r}
pheno_filt$Description

to_remove <- c(
  # Not specific
  "Cancer, suspected or other",
  "Secondary malignant neoplasm",
  "Neoplasm of uncertain behavior",
  "Lipoma",
  "Vascular hamartomas and non-neoplastic nevi",
  "Nevus, non-neoplastic",
  # Endocrine
  "Nontoxic nodular goiter",
  "Nontoxic uninodular goiter",
  "Hypothyroidism",
  "Hypothyroidism NOS",
  "Type 2 diabetes with renal manifestations",
  "Type 2 diabetes with ophthalmic manifestations",
  "Type 2 diabetes with neurological manifestations",
  "Insulin pump user",
  "Other abnormal glucose",
  "Polyneuropathy in diabetes",
  "Diabetic retinopathy",
  # Unrelated
  "Gout and other crystal arthrop",
  "Gout",
  "Back pain",
  "Thoracic or lumbosacral neuritis or radiculitis, unspecified",
  "Sciatica",
  "Neuralgia, neuritis, and radiculitis NOS",
  "Nonallopathic lesions NEC",
  "Myalgia and myositis unspecified",
  "Edema",
  "Fever of unknown origin",
  "Syncope and collapse",
  "Nausea and vomiting",
  "Other abnormal blood chemistry",
  "Malaise and fatigue",
  "Debility unspecified",
  # Digestive, but not specific or overlaps with another code
  "Overweight, obesity and other hyperalimentation",
  "Diseases of esophagus",
  "Esophagitis, GERD and related diseases",
  "Symptoms involving digestive system",
  "Diverticulosis and diverticulitis",
  "Anal and rectal conditions",
  "Abnormal results of function study of liver"
  # Unrelated biomarkers
)

pheno_filt <- pheno_filt %>% filter(!Description %in% to_remove)
```


# Loading in file manifests
```{r}
# Three batches of lab data
lab_1 <- read.table("https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/analyses/GIA/phs002453.MVP_R4.1000G_AGR.GIA.Labs_batch1.analysis-PI.MULTI.tar.table_of_contents.txt")
lab_2 <- read.table("https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/analyses/GIA/phs002453.MVP_R4.1000G_AGR.GIA.Labs_batch2.analysis-PI.MULTI.tar.table_of_contents.txt")
lab_3 <- read.table("https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/analyses/GIA/phs002453.MVP_R4.1000G_AGR.GIA.Labs_batch3.analysis-PI.MULTI.tar.table_of_contents.txt")

# PheCode digestive system
digest_1 <- read.table("https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/analyses/GIA/phs002453.MVP_R4.1000G_AGR.GIA.PheCodes_DigestiveSystem_batch1.analysis-PI.MULTI.tar.table_of_contents.txt")
digest_2 <- read.table("https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/analyses/GIA/phs002453.MVP_R4.1000G_AGR.GIA.PheCodes_DigestiveSystem_batch2.analysis-PI.MULTI.tar.table_of_contents.txt")

# PheCode endocrine or metabolic
endo_1 <- read.table("https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/analyses/GIA/phs002453.MVP_R4.1000G_AGR.GIA.PheCodes_EndocrineMetabolic_batch1.analysis-PI.MULTI.tar.table_of_contents.txt")
endo_2 <- read.table("https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/analyses/GIA/phs002453.MVP_R4.1000G_AGR.GIA.PheCodes_EndocrineMetabolic_batch2.analysis-PI.MULTI.tar.table_of_contents.txt")

# PheCode neoplasms
neo_1 <- read.table("https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/analyses/GIA/phs002453.MVP_R4.1000G_AGR.GIA.PheCodes_Neoplasms_batch1.analysis-PI.MULTI.tar.table_of_contents.txt")
neo_2 <- read.table("https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/analyses/GIA/phs002453.MVP_R4.1000G_AGR.GIA.PheCodes_Neoplasms_batch2.analysis-PI.MULTI.tar.table_of_contents.txt")

#PheCode symptoms
symptoms_1 <- read.table("https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/analyses/GIA/phs002453.MVP_R4.1000G_AGR.GIA.PheCodes_Symptoms.analysis-PI.MULTI.tar.table_of_contents.txt")

# Vitals
vitals_1 <- read.table("https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/analyses/GIA/phs002453.MVP_R4.1000G_AGR.GIA.Surveys_Vitals_Quantitative.analysis-PI.MULTI.tar.table_of_contents.txt")

```

# Merging and organizing the tables of contents
```{r}
file_names <- rbind(lab_1, lab_2, lab_3,
                   digest_1, digest_2,
                   endo_1, endo_2, 
                   neo_1, neo_2,
                   symptoms_1, vitals_1)

rm(lab_1, lab_2, lab_3,
   digest_1, digest_2,
   endo_1, endo_2,
   neo_1, neo_2,
   symptoms_1, vitals_1)

# Filter for European ancestry file names
file_names <- file_names %>% filter(str_detect(V6, "EUR"))

# Pull trait name out of file name for column merging
file_names <- file_names %>% mutate(Trait = sub(".*?AGR\\.(.*?)\\.EUR.*", "\\1", V6))

# Join tables
pheno_filt <- left_join(pheno_filt, file_names)

```

# Pulling out tarball names

```{r}
pheno_filt <- pheno_filt %>% mutate(tar_file = "phs002453." %>% paste0(str_extract(V6, "^[^/]+")) %>% paste0(".analysis-PI.MULTI.tar"))

unique_tarballs <- unique(pheno_filt$tar_file)
unique_tarballs

unique_md5s <- paste0(unique_tarballs, ".md5")
unique_md5s

pheno_filt <- pheno_filt %>% mutate(file_name = gsub(".*/", "", V6))

```

# Test downloading a small tarball to the server
```{r}
#system("wget -P /dartfs/rc/lab/S/Szhao/katieh/organoid-rotation/data/ https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/analyses/GIA/phs002453.MVP_R4.1000G_AGR.GIA.PheCodes_CongenitalAnomalies.analysis-PI.MULTI.tar")
#system("wget -P /dartfs/rc/lab/S/Szhao/katieh/organoid-rotation/data/ https://ftp-ncbi-nlm-nih-gov.dartmouth.idm.oclc.org/dbgap/studies/phs002453/phs002453.v1.p1/analyses/GIA/phs002453.MVP_R4.1000G_AGR.GIA.PheCodes_CongenitalAnomalies.analysis-PI.MULTI.tar.md5")

```


# Downloading files to the server
```{r}
write.csv(pheno_filt, "/dartfs/rc/lab/S/Szhao/katieh/organoid-rotation/data/MVP_pheno_list.csv")

```

Files are then downloaded and extracted one-by-one using the MVP-data-pull pipeline found in code/pipeline/MVP-data-pull/

