---
title: "UK Biobank Phenotypes"
author: "Katie Hooker"
date: "2025-01-15"
output: html_document
---

# Reading in the [phenotype manifest](https://pan.ukbb.broadinstitute.org/docs/per-phenotype-files)

```{r}
library(readxl)

pheno_manifest <- read.csv("/dartfs/rc/lab/S/Szhao/katieh/organoid-rotation/data/UKBB/ukbb_phenotype_manifest.csv")

h2_manifest <- read.csv("/dartfs/rc/lab/S/Szhao/katieh/organoid-rotation/data/UKBB/ukbb_h2_manifest.csv")

```

# APPROACH 1: Restricting to 10K people in the cohort, followed by keyword restrictions
This approach filters for 10k cohort size, keywords, and then European heritability greater than or equal to 0.1.

## Restricting to full cohort size \>= 10,000

```{r}
library(dplyr)

TenK_pheno <- subset(pheno_manifest, n_cases_full_cohort_both_sexes >= 10000)

```


## Restricting phenotypes by keyword(s)

```{r}
library(stringr)

# Searching for keywords
TenK_pheno_filtered <- TenK_pheno %>% filter(
  str_detect(category, "Biological samples") |
    str_detect(category, "Chapter IV Endocrine, nutritional and metabolic diseases") |
    str_detect(category, "Chapter XI Diseases of the digestive system") |
    str_detect(category, "R10-R19 Symptoms and signs involving the digestive system and abdomen") |
    category == "endocrine/metabolic" |
    category == "digestive" |
    category == "neoplasms" |
    category == "congenital anomalies"|
    str_detect(category, "D10-D36 Benign neoplasms") |
    str_detect(trait_efo_categories, "Digestive system") |
    str_detect(trait_efo_categories, "Lipid or lipoprotein measurement") |
    str_detect(trait_efo_categories, "Liver enzyme measurement") |
    str_detect(trait_efo_categories, "Metabolic disorder"))
    # str_detect(trait_efo_categories, "Cancer") |
    # str_detect(trait_efo_categories, "Other disease") |
    # str_detect(trait_efo_categories, "Other measurement") |
    # str_detect(trait_efo_categories, "Other trait"))

print(nrow(TenK_pheno_filtered))

```


```{r}
# Removing diseases of family members
TenK_pheno_filtered <- TenK_pheno_filtered %>% filter(!startsWith(description, "Illnesses of"))
print(nrow(TenK_pheno_filtered))

# Removing blood count assays
TenK_pheno_filtered <- TenK_pheno_filtered %>% filter(str_detect(category, "(?i)Blood count", negate=TRUE)) # (?i) specifies to ignore case
print(nrow(TenK_pheno_filtered))

# Removing abnormalities in urine
TenK_pheno_filtered <- TenK_pheno_filtered %>% filter(str_detect(description, "(?i)in urine", negate=TRUE))
print(nrow(TenK_pheno_filtered))

# Removing measures adjusted by medication
TenK_pheno_filtered <- TenK_pheno_filtered %>% filter(str_detect(description, "(?i)adjusted by medication", negate=TRUE))
print(nrow(TenK_pheno_filtered))

# Removing other non-digestive traits
TenK_pheno_filtered <- TenK_pheno_filtered %>% 
  filter(str_detect(description, "(?i)Breast", negate=TRUE) &
           str_detect(description, "(?i)skin", negate=TRUE) &
           str_detect(description, "(?i)uter", negate=TRUE))
print(nrow(TenK_pheno_filtered))

# Removing non-digestive specific descriptions
TenK_pheno_filtered <- TenK_pheno_filtered %>% 
  filter(description != "Cancer, suspected or other" &
  description != "Malignant neoplasm, other" &
  description != "Chemotherapy" &
  description != "Secondary malignant neoplasm")
print(nrow(TenK_pheno_filtered))

```

## Restricting to European $h^2 >= 0.1$

```{r}
EUR_h2 <- h2_manifest[h2_manifest$pop == "EUR",]
EUR_h2 <- EUR_h2 %>% select(trait_type, phenocode, pheno_sex, coding, modifier, pop,
                            estimates.ldsc.h2_observed)
EUR_h2$phenocode <- as.character(EUR_h2$phenocode)
# Changing h2 to numeric
EUR_h2 <- EUR_h2 %>% mutate(h2_EUR_numeric = as.numeric(estimates.ldsc.h2_observed))

TenK_pheno_filtered <- left_join(TenK_pheno_filtered, EUR_h2, by = c("trait_type","phenocode","pheno_sex","coding","modifier")) 

# Subsetting for EUR h2 >= 0.1
TenK_pheno_filtered <- TenK_pheno_filtered %>% subset(h2_EUR_numeric >= 0.1)
print(nrow(TenK_pheno_filtered))

```

## Final organizing

```{r}
#TenK_pheno_filtered$description
#TenK_pheno_filtered$coding_description

final_pheno <- select(TenK_pheno_filtered, trait_type, phenocode, pheno_sex,
                     coding, modifier, description, description_more, coding_description,
                     category, n_cases_full_cohort_both_sexes, pops_pass_qc, estimates.ldsc.h2_observed, 
                     aws_link)
final_pheno
```

# APPROACH 2: Keyword restrictions only
This approach filters for keywords and then European heritability greater than or equal to 0.1 (no cohort size restriction).

## Restricting phenotypes by keyword(s)

```{r}
library(stringr)

# Searching for keywords
pheno_filtered <- pheno_manifest %>% filter(
    str_detect( category, "Biological samples") |
    str_detect(category, "Chapter IV Endocrine, nutritional and metabolic diseases") |
    str_detect(category, "Chapter XI Diseases of the digestive system") |
    str_detect(category, "R10-R19 Symptoms and signs involving the digestive system and abdomen") |
    category == "endocrine/metabolic" |
    category == "digestive" |
    category == "neoplasms" |
    category == "congenital anomalies"|
    str_detect(category, "D10-D36 Benign neoplasms") |
    str_detect(trait_efo_categories, "Digestive system") |
    str_detect(trait_efo_categories, "Lipid or lipoprotein measurement") |
    str_detect(trait_efo_categories, "Liver enzyme measurement") |
    str_detect(trait_efo_categories, "Metabolic disorder"))
    # str_detect(trait_efo_categories, "Cancer") |
    # str_detect(trait_efo_categories, "Other disease") |
    # str_detect(trait_efo_categories, "Other measurement") |
    # str_detect(trait_efo_categories, "Other trait"))

print(nrow(pheno_filtered))

```


```{r}
# Removing diseases of family members
pheno_filtered <- pheno_filtered %>% filter(!startsWith(description, "Illnesses of"))
print(nrow(pheno_filtered))

# Removing blood count assays
pheno_filtered <- pheno_filtered %>% filter(str_detect(category, "(?i)Blood count", negate=TRUE)) # (?i) specifies to ignore case
print(nrow(pheno_filtered))

# Removing abnormalities in urine
pheno_filtered <- pheno_filtered %>% filter(str_detect(description, "(?i)in urine", negate=TRUE))
print(nrow(pheno_filtered))

# Removing measures adjusted by medication
pheno_filtered <- pheno_filtered %>% filter(str_detect(description, "(?i)adjusted by medication", negate=TRUE))
print(nrow(pheno_filtered))

# Removing other non-digestive traits
pheno_filtered <- pheno_filtered %>% 
  filter(str_detect(description, "(?i)Breast", negate=TRUE) &
           str_detect(description, "(?i)skin", negate=TRUE) &
           str_detect(description, "(?i)uter", negate=TRUE))
print(nrow(pheno_filtered))

# Removing non-digestive specific descriptions
pheno_filtered <- pheno_filtered %>% 
  filter(description != "Cancer, suspected or other" &
  description != "Malignant neoplasm, other" &
  description != "Chemotherapy" &
  description != "Secondary malignant neoplasm")
print(nrow(pheno_filtered))

```

## Restricting to European $h^2 >= 0.1$

```{r}
pheno_filtered <- left_join(pheno_filtered, EUR_h2, by = c("trait_type","phenocode","pheno_sex","coding","modifier")) 

# Subsetting for EUR h2 >= 0.1
pheno_filtered <- pheno_filtered %>% subset(h2_EUR_numeric >= 0.1)
print(nrow(pheno_filtered))

```

## Checking to see how many TOTAL traits have EUR $h2 >= 0.1$
```{r}
EUR_h2_filtered <- EUR_h2 %>% subset(h2_EUR_numeric >= 0.1)

EUR_h2_filtered$phenocode <- as.character(EUR_h2_filtered$phenocode)
pheno_manifest$phenocode <- as.character(pheno_manifest$phenocode)

pheno_details <- pheno_manifest[c(1:10),]
test <- left_join(EUR_h2_filtered, pheno_details)

nrow(test)
```


## Final organizing

```{r}
#pheno_filtered$description
#pheno_filtered$coding_description

final_pheno2 <- select(pheno_filtered, trait_type, phenocode, pheno_sex,
                     coding, modifier, description, description_more, coding_description,
                     category, n_cases_full_cohort_both_sexes, pops_pass_qc, estimates.ldsc.h2_observed)
final_pheno2
```

Multiple filtering techniques lead us to only 22 biomarker phenotypes of relevance due to low heritability. As a result, we are pivoting to using the All of Us dataset instead of the UK biobank.

# VALIDATION OF MVP RESULTS
We are interested in validating some enrichment trends observed in the MVP GWAS enrichment analysis. As such, we will perform enrichment analysis on the 22 biomarker phenotypes from UK biobank in addition to LDL cholesterol and glucose. These last two traits were originally filtered out due to heritability < 0.1.

## Adding back in the two traits of interest
```{r}

two_phenos <- pheno_manifest %>% subset(description == "LDL direct"|
                                            description == "Glucose") %>% 
  left_join(EUR_h2) %>% 
  select(trait_type, phenocode, pheno_sex,
         coding, modifier, description, description_more, coding_description,
         category, n_cases_full_cohort_both_sexes, pops_pass_qc, estimates.ldsc.h2_observed,
         aws_link)

validation_pheno <- full_join(final_pheno, two_phenos)

```
# Downloading the summary statistics
GWAS summary statistics were downloaded from IEU OpenGWAS based on phenocode.

```{r}

validation_pheno <- validation_pheno %>% mutate(file_name = paste0("ukb-d-", phenocode, "_irnt"))

file_list <- validation_pheno$file_name

# for (file_name in file_list){
#   system(paste0("wget -P /dartfs/rc/lab/S/Szhao/katieh/organoid-rotation/raw-sumstats/UKBB/ https://gwas.mrcieu.ac.uk/files/", file_name, "/", file_name, ".vcf.gz"))
# }


```

# Saving phenotype list
```{r}
write.csv(validation_pheno, "/dartfs/rc/lab/S/Szhao/katieh/organoid-rotation/data/UKBB_pheno_list_validation.csv")
```

